\LARGE \bf Artificial Cognition for Social Human-Robot Interaction:\\ An Implementation

### 2016-07-04 paper.tex
Second round of proof-reading 4

The previous sections provide a perspective on a complete deliberative
architecture for social robots, including an implementation supported
``real-world'' experimental deployments. This section synthesises next what we
see as the main challenges that human-robot interaction brings to Artificial
Intelligence.  We discuss first how embodied cognition is an essential challenge
in human-robot interaction; we rephrase then the requirements of joint actions
in terms of five questions; we discuss the importance of building and
maintaining a multi-level model of the human; and we finally reflect on the
importance of explicit knowledge management in robotic architectures that deal
with human-level semantics and state in that respect the current limits of our
logic framework.

@@

Robotics is traditionally regarded as the prototypical instance of
\emph{embodied} artificial intelligence, and this dimension is especially
prevalent in human-robot interaction, where the robot has to share and
collaborate in a joint physical environment.  This leads to a tight coupling
between the symbolic and the geometric realms: while AI at its origins was
mostly concerned with symbolic models, it has been since recognised that not
only the mind is not a purely abstract system, disconnected from the physical
world, but even more, cognition fundamentally relies on its relation to the
physical world (so-called \emph{embodied cognition}). Varela~\cite{Varela1992}
is one of the main discoverers of these mechanisms, and coined the concept of
\emph{enactivism} as the theoretical framework that studies the links between
cognition, embodiment and actions.  This has since been thoroughly studied in
robotics and artificial intelligence (Pfeifer and
Bongard~\cite{pfeifer2007body} is one of the reference).

@@

Grounding also occurs during verbal and non-verbal communication. The {\sc
Dialogs} module grounds new concepts introduced by the speaker by asking
questions until it can attach the new concept to concepts already present in the
knowledge pool (if one asks the robot ``bring me a margarita'', the robot may
initially asks ``what is a margarita?''. The user would answer ``a cocktail''
and the robot would continue the grounding -- ``what is a cocktail?'' -- until
it anchors the new concept to ones it already knows). Embodied interactions
(like gestures) are also taken into account at this level: we have presented how
pointing, for instance, is used by the robot to ground \emph{``this''} or
\emph{``that''} to the pointed artifact.

Note that, because only objects marked with 2D fiducial markers are currently
recognised (typically, about ten of them are simultaneously used in a given
experiment), our grounding mechanisms have only been exercised in small-sized
closed world. This simplifies the task, and we can not claim that our approach
provides a generic grounding capability. Beyond the perception of geometric
entities, context, cultural background, ``naive physics'' knowledge, emotional
state of the human are some of the numerous other determinants that need to be
accounted for when grounding human-robot interaction.

@@

Our context, where a robot has to achieve a task together with
humans, raises issues to be tackled by the robots' decisional components, and
that we can summarise as ``W-questions'': \emph{What}, \emph{Who}, \emph{When},
\emph{Where} and \emph{How}?

@@

\emph{When to perform a given action} is enriched here by the human context
since it has to take into account the human, her needs, her rhythms and
pace, and her mental state. While performing its share of the task, the
robot has to produce signals directed towards the human and to respond to
signals produced back at the proper pace.

@@

\emph{How to perform an action}, finally, needs to be reflected upon by the
robot: several options to perform an action or to achieve a goal are often
available, and selecting one is a non-trivial decision problem. Cost-based
planners are one possible approach: they search for plans that satisfy an
acceptable cost in terms of \emph{acceptability} or \emph{legibility}, for
instance.

@@

They are in fact considered in each of the components, at different levels of
abstraction (from the abstract shared plan level to the action refinement and
execution level).  At every interaction step, the choice of the next action, of
who has to act (the human, the robot or both of them) and when to perform it is
made by {\sc Shary} with the help of HATP, based on perspective-taking and
situation assessment ({\sc Spark} and {\sc Oro} as well as on estimation of the
human mental state ({\sc Shary} with the help of {\sc Oro}). The actual
realization of the action is eventually supported by a full set of on-line
planning functions provided by {\sc Mhp} that computes not only trajectories but
also the pertinent placements and postures based not only on geometric
information but also estimation of the human current mental state and
preferences ({\sc Oro}).

@@

As thoroughly presented in this article, we have built the decisional
capabilities of our robots around this idea of explicit knowledge manipulation.
We finally briefly comment on this design choice and the limitations of our
approach.

\subsubsection{Explicit Knowledge in our Architecture}

The components that we have presented so far build a \emph{knowledge-oriented}
architecture: knowledge is explicitly stored in one central and consistent
repository of facts, accessible to all modules. It relies on a strict formalism
(OWL statements), with a well defined vocabulary (stated in the common-sense
ontology). These first two points lead to a loosely-coupled architecture
where modules can be removed or replaced easily by other ones as long as they
share the same semantics: modules are defined by the knowledge they produce or
consume.

@@

Those characteristics are not novel \emph{per se}. We want however underline the
shift of focus brought by this approach during the design and integration phases
of robots: components of our deliberative layer are defined and bound together
by the knowledge they produce and consume. Human-robot interaction, because it
supposes operations at \emph{human-level} and in environments with complex
semantics, acts here as a motivational force.

\subsubsection{Limits of Disambiguation at Semantic Level}

@@

When playing this kind of game, however, the issue arises that the robot has no
way to select which knowledge about the object is relevant in the interaction
context. For instance, the knowledge base may store facts like \stmt{OBJ1 type
ActiveConcept} (which internally means that this concept was mentioned in a
discussion in the last few seconds): this information is not a relevant
property of \concept{obj1} when trying to disambiguate concepts with humans.
This distinction between \emph{internal knowledge} (meaningful to
the system only) and \emph{common knowledge} (whose meaning is understood by
all the interactors) has not been properly dealt with in our architecture.

Besides, even knowledge that belongs to the \emph{common knowledge} may not be
appropriate in a given interaction context. For instance, the system may
compute that at a given instant the human is looking at the object: \stmt{HUMAN
looksAt obj1}. This property makes sense to both parties, but in the context of
the \emph{spygame}, we would like to mainly use immanent properties, not
volatile like a gaze. More research is required to identify relevant
interaction contexts and knowledge classes attached to them.

@@

The integration of these components in a consistent, working and observable
system builds upon the particular design of the interfaces between the cognitive
components: the information streams use high-level semantics, represented as
first-order logic statements. In that sense, our deliberative architecture is
similar to projects like {\sc cram}~\cite{Beetz2010}, KeJia~\cite{Chen2010} or
PEIS Ecology~\cite{saffiotti2005peis, daoutis2012cooperative}, with however a
stronger emphasise on the representation and interaction with humans. We however
distinguish ourselves from research on \emph{cognitive architectures}: cognitive
architectures are usually understood as an artificial yet principled model of
(human) cognition. While some have been deployed on autonomous robots, like
HAMMER~\cite{demiris2006hierarchical} or ACT-R/E~\cite{trafton2013act}, most are
primarily concerned with the modeling of human cognition and are less focused on
practical deployment on socially interactive robots.

In that sense, our contribution in terms of architecture is a practical one:
our integration model enables to consistently combine a large set of
technically independent yet cognitively interdependent cognitive processes. We
bridge them through explicit, human-level semantics, and we show that this
results in a fully implemented system, effectively deployed on several
platforms and in several real interaction scenarios.

Besides, this work introduces several new contributions related to the
representation and the management of humans in an autonomous robotic system.
Specifically, we mentioned in the introduction the following four points:

@@

While several contributions in the literature provide insights and
contributions on one aspect or another (references are in the corresponding
subsections), we are not aware of a fully implemented architecture that
effectively combines in a coherent manner all these points. Our approach
allowed us to identify, refine and combine a number of key cognitive abilities
that we believe are needed for successful and autonomous social human-robot
interaction. The novelty and relevance of this contribution to HRI is
underlined by the range of multi-disciplinary collaborations and studies that
have been made possible by our
architecture~\cite{dautenhahn2006may,koay2007exploratory,Ros2010b,dehais2011physiological,ferreira2015users,GharbiROMAN2015}.

@@

There is also extensive work to be done in order to refine the notion of ``good
shared plan'' and ``good/acceptable robot behaviour'' in this context. There are
obviously large avenues for learning and adaptation in this context.

Another direction to head to deals with \emph{contexts representation}.
Contexts are currently often limited to the current spatial and temporal
situation. Some models offer the possibility to jump in the past or to switch to
another agent's perspective, but in current approaches, selecting a context
always basically consists in retrieving a set of beliefs corresponding to a
situation, and temporarily replacing the current beliefs by those other ones.
This misses the fact that at a given moment, not one but many contexts co-exist
at different scales. We do not want to retrieve one monolithic set of beliefs,
but instead carefully craft a context from several \emph{atomic} contexts.
Techniques for representation of overlapping pools of knowledge largely remain
to be developed, as well as efficient algorithms to retrieve (or discard) such
context-related pools of knowledge. This is a challenge not only for robotics,
but more generally for artificial intelligence.  The ability to explicitly
manage contexts and context switches would endow the robot with a cognitive
capability similar to what is known as \emph{context-dependent memory} in
cognitive psychology. This is also related to Tulving's \emph{autonoetic
consciousness}~\cite{Tulving1985a}: the ability to reflect upon its own past or
future experiences.  Much remain to be done to this regard, starting with a
formal analysis of what are the relevant contexts for our robots.

Human-Robot Interaction is and is going to remain a challenging field for
Artificial Intelligence. We hope that this contribution helps with clarifying
these challenges and making them concrete decisional problems.

>>>>

The previous sections provide a perspective on a complete deliberative
architecture for social robots, including its implementation, supported
by experimental deployments. This section synthesises what we
see as the main challenges that human-robot interaction brings to Artificial
Intelligence.  We first discuss how embodied cognition is an essential challenge
in human-robot interaction; we rephrase then the requirements of joint actions
in terms of five questions; we discuss the importance of building and
maintaining a multi-level model of the human; and we finally reflect on the
importance of explicit knowledge management in robotic architectures that deal
with human-level semantics and state in that respect the current limits of our
logic framework.

@@

Robotics is traditionally regarded as the prototypical instance of
\emph{embodied} artificial intelligence, and this dimension is especially
prevalent in human-robot interaction, where the robot has to share and
collaborate in a joint physical environment.  This leads to a tight coupling
between the symbolic and the geometric realms: while AI at its origins was
mostly concerned with symbolic models, it has been since recognised that not
only the mind is not a purely abstract system, disconnected from the physical
world, but even more, cognition fundamentally relies on its relation to the
physical world (so-called \emph{embodied cognition}). Varela~\cite{Varela1992}
is one of the main discoverer of these mechanisms, and coined the concept of
\emph{enactivism} as the theoretical framework that studies the links between
cognition, embodiment and actions.  This has since been thoroughly studied in
robotics and artificial intelligence (Pfeifer and
Bongard~\cite{pfeifer2007body} is one of the reference).

@@

Grounding also occurs during verbal and non-verbal communication. The {\sc
Dialogs} module grounds new concepts introduced by the speaker by asking
questions until it can attach the new concept to concepts already present in the
knowledge pool (if one asks the robot ``bring me a margarita'', the robot may
initially asks ``what is a margarita?''. The user would answer ``a cocktail''
and the robot would continue the grounding -- ``what is a cocktail?'' -- until
it anchors the new concept to ones it already knows like \concept{Drink}). Embodied interactions
(like gestures) are also taken into account at this level: we have presented how
pointing, for instance, is used by the robot to ground \emph{``this''} or
\emph{``that''} to the pointed artifact.

Note that, because only objects marked with 2D fiducial markers are currently
recognised (typically, about ten of them are simultaneously used in a given
experiment), our grounding mechanisms have only been exercised in small-sized
closed world. This simplifies the task, and we can not claim that our approach
provides a generic grounding capability. Context, cultural background, ``naive physics'' knowledge, emotional
state of the human are some of the numerous determinants beyond the perception of geometric
entities. They would need to be accounted for when grounding human-robot interaction.

@@

Our context, where a robot has to achieve a task together with
humans, raises another set of specific issues to be tackled by the robot's
decisional components. We summarise them as the ``W-questions'': \emph{What}, \emph{Who}, \emph{When},
\emph{Where} and \emph{How}?

@@

\emph{When to perform a given action} is made more complex by the presence of
humans. The robot has to take into account her needs, her rhythms and
pace, and her mental state. While performing its share of the task, the
robot has to produce signals directed towards the human and to respond to
signals produced back at the proper pace to ensure collaboration.

@@

\emph{How to perform an action}, finally, needs to be reflected upon by the
robot: several options to perform an action or to achieve a goal are often
available, and selecting one is a non-trivial decision problem. Cost-based
planners augmented with social rules are one possible approach: they search for plans that satisfy an
acceptable cost in terms of \emph{acceptability} or \emph{legibility} as well.

@@

They are in fact considered in each of the components, at different levels of
abstraction (from the abstract shared plan level to the action refinement and
execution level).  At every interaction step, the choice of the next action, of
who has to act (the human, the robot or both of them) and when to perform it is
made by {\sc Shary} with the help of HATP, based on perspective-taking and
situation assessment ({\sc Spark} and {\sc Oro}), as well as on the estimation of the
human mental state ({\sc Shary} with the help of {\sc Oro}). The actual
realization of the action is eventually supported by a full set of on-line
planning functions provided by {\sc Mhp} that computes not only trajectories but
also the pertinent placements and postures based not only on geometric
information but also estimation of the human current mental state and
preferences ({\sc Oro}).

@@

As thoroughly presented in this article, we have built the decisional
capabilities of our robots around this idea of explicit knowledge manipulation.

\paragraph{Explicit knowledge in our architecture}

The components that we have presented so far build a \emph{knowledge-oriented}
architecture: knowledge is explicitly stored in one central and consistent
repository of facts, accessible to all modules. It relies on a strict formalism
(OWL statements), with a well defined vocabulary (stated in the common-sense
ontology). These first two points lead to a loosely-coupled architecture
where modules can be removed or replaced by other ones as long as they
share the same semantics: modules are defined by the knowledge they produce or
consume.

@@

We want to underline the
shift of focus brought by this approach during the design and integration phases
of robots: components of our deliberative layer are defined and bound together
by the knowledge they produce and consume. Human-robot interaction, because it
supposes operations at human level and in environments with complex
semantics, acts here as a motivational force.

\paragraph{Limits of disambiguation at semantic level}

@@

When playing this kind of game, however, the issue arises that the robot has no
way to select which knowledge about the object is relevant in the interaction
context. For instance, the knowledge base may store facts like \stmt{OBJ1 type
ActiveConcept} (which internally means that this concept was mentioned in a
discussion in the last few seconds): this information is not a relevant
property of \concept{OBJ1} when trying to disambiguate concepts with humans.
This distinction between \emph{internal knowledge} (meaningful to
the system only) and \emph{common knowledge} (whose meaning is understood by
all the agents) has not been properly dealt with in our architecture.

Besides, even knowledge that belongs to the \emph{common knowledge} may not be
appropriate in a given interaction context. For instance, the system may
compute that at a given instant the human is looking at the object: \stmt{HUMAN
looksAt OBJ1}. This property makes sense to both parties, but in the context of
the \emph{spygame}, we would like to mainly use immanent properties, not
volatile like a gaze. More research is required to identify relevant
interaction contexts and knowledge classes attached to them.

@@

The integration of these components in a consistent, working and observable
system builds upon the particular design of the interfaces between the cognitive
components: the information streams use high-level semantics, represented as
first-order logic statements. In that sense, our deliberative architecture is
similar to projects like {\sc cram}~\cite{Beetz2010}, KeJia~\cite{Chen2010} or
PEIS Ecology~\cite{saffiotti2005peis, daoutis2012cooperative}, with however a
stronger emphasise on the specificities of the interaction with humans.
Importantly, we
distinguish ourselves from research on \emph{cognitive architectures}: cognitive
architectures are usually understood as an artificial yet principled model of
(human) cognition. While some have been deployed on autonomous robots, like
HAMMER~\cite{demiris2006hierarchical} or ACT-R/E~\cite{trafton2013act}, most are
primarily concerned with the modeling of human cognition and are less focused on
the effective deployment on socially interactive robots.
In that sense, our contribution in terms of architecture is a practical one:
our integration model enables to consistently combine a large set of
technically independent yet cognitively interdependent cognitive processes. We
bridge them through explicit, human-level semantics, and we show that this
results in a fully implemented system, effectively deployed on several
platforms and in several real interaction scenarios.

This work introduces several new contributions related to the
representation and the management of humans in an autonomous robotic system.
Specifically, we mentioned in the introduction the following four points:

@@

While several contributions in the literature provide insights and
contributions on one aspect or another (references are in the corresponding
subsections), we are not aware of a fully implemented architecture that
effectively combines in a coherent manner all these points. The novelty and relevance of this contribution to HRI is
further underlined by the range of multi-disciplinary collaborations and studies that
have been made possible by our
architecture~\cite{dautenhahn2006may,koay2007exploratory,Ros2010b,dehais2011physiological,ferreira2015users,GharbiROMAN2015}.

@@

There is also extensive work to be done in order to refine the notion of ``good
shared plan'' and ``good/acceptable robot behaviour'' in this context. There are
large avenues for learning and adaptation in this context.

Another direction to head to deals with context representation.
Contexts are currently often limited to the current spatial and temporal
situation. Some of our models offer the possibility to jump in the past or to switch to
another agent's perspective, but in our current approach, selecting a context
essentially consists in retrieving a set of beliefs corresponding to a
situation, and temporarily replacing the current beliefs by those other ones.
This misses the fact that at a given moment, not one but many contexts co-exist
at different scales. We do not want to retrieve one monolithic set of beliefs,
but instead carefully craft a context from several \emph{atomic} contexts.
Techniques for representation of overlapping ``pools'' of knowledge largely remain
to be developed, as well as efficient algorithms to retrieve (or discard) such
context-related pools of knowledge. This is a challenge not only for robotics,
but more generally for artificial intelligence.  The ability to explicitly
manage contexts and context switches would endow the robot with a cognitive
capability similar to what is known as \emph{context-dependent memory} in
cognitive psychology. This is also related to Tulving's \emph{autonoetic
consciousness}~\cite{Tulving1985a}: the ability to reflect upon its own past or
future experiences.  Much remain to be done to this regard, starting with a
formal analysis of what are the relevant contexts for our robots.

Human-Robot Interaction is and will remain a challenging field for
Artificial Intelligence. We hope that this contribution helps with clarifying
these challenges and making them concrete decisional problems.
